\specialhead{ACKNOWLEDGMENT}

\noindent This dissertation is a product of patience.  The highest amounts of mental and physical endurance have been given by my wife, Lindsay.  Her skill in raising our children, Heidi and Lincoln, has been awe-inspiring, and I thank her sincerely for her perseverance.  It can be challenging to explain to young children why work can be so important -- important enough to miss dinners or playtimes at the park.  I believe and hope that someday they will understand that the missed time together was instead invested into a higher, more urgent obligation.  I dedicate this work to them, as they will be the truest beneficiaries of any success my work may find in the pursuit of wildlife conservation.

I also thank my advisor, Dr. Charles Stewart, who has been a compassionate guide in my academic career and in life as a young husband and father.  I appreciate his patience and expertise and the guidance from my committee.  The generosity and flexibility of my employers during this Ph.D.\ process is something that I'm not sure I completely understand; I thank Drs. Anthony Hoogs, Matt Turek, Rusty Blue, and Keith Fieldhouse at Kitware, along with Jason Holmberg and Dr. Tanya Berger-Wolf with Wild Me.  I also thank the Gordon and Betty Moore Foundation for their financial support.  My graduate lab partners at RPI, Drs. Jon Crall and Hendrik Weideman, provided excellent discussions and stimulation on the latest machine learning methods, and I thank Dr. Barbara Cutler for her tranquility in indulging our energy.  Jon, if you ever want to camp in the African bush, just let me know.  I also thank my peers in machine learning for animal conservation, Sara Beery and Dr. Stephan Schneider, for their work in kindling a small but passionate research community.  Drs. Dan Rubenstein, Kaia Tombak, and Megan McSherry have also been instrumental in facilitating this research, and I thank them for their diligence in working with me over the years.  I also cannot forget the dedication and benevolence of the research staff at the Ol Pejeta and Lewa conservancies, the Great Gr\'evy's Trust, the Kenya Wildlife Service, and numerous Wildbook projects.

Lastly, I would like to thank my parents, Anthony, Grace, Linda, Harlon Jr., Kent, and Julie, and my siblings Stephany, Harlon III, Joyce, Kelsey, Kyle, and Chad for their continued support.  I also appreciate my co-workers at Wild Me, Jon Van Oast, Drew Blount, Colin Kingen, Mark Fisher, Ben Schiener, and Tanya Stere for permitting my chaos and giving me a fulfilling place to work with friends.  I also thank Drew and Olga Moskvyak for their work on new detection components and Tanya as honorary editor.  Specific thanks to my sisters-in-law Brittany and Kelsey Sundman, and to Ben and Kaia, for their last-minute help, looking at some zebras when nobody else really wanted to.
